Convergence rate of MAP estimates \\ for the exponential family

### 2021-10-13 aistats.tex
Add ref to large deviations

Exponential families are among the most widely used statistical models.
Many standard random variables fall into this category: Gaussians, categorical, Gamma or Dirichlet.
They are both flexible enough to model a variety of data sources $X$, and easy to describe with some sufficient statistics $T(X) \in \real^d$.
In particular, they are appreciated for their convex log-likelihood
\alignn{
f(\nat) = \E[-\log p(X \cond \nat)] = \logpart(\nat) - \lin{\E[T(X)] , \nat}
}
where $\logpart$ is a convex function and $\nat$ is a parameter vector.
This convexity lays the foundation for generalized linear models \citep{mccullagh1989generalized} 
or variants of principal component analysis \citep{collins2001generalization} among other applications.
Another interest of exponential families is their closed form maximum-likelihood estimate (MLE) obtained via moment matching
\begin{align}
	\nabla \logpart(\hat \nat_n^\text{MLE}) = \frac{\sum_{i=1}^n T(x_i)}{n} 
\end{align}
Given a conjugate prior, a similar formula holds for the maximum a posteriori (MAP).

@@

\paragraph{Maximum A Posteriori (MAP).} 
\textcolor{red}{\textbf{D.S.: This section is very unclear. }}
Using Bayes rule with~\eqref{eq:def_expfamily}~\&~\eqref{eq:def_prior}, the posterior of $\nat$ given a dataset $\mD_n =(X_1,\dots,X_n)$ is
\begin{align}
	p(\nat \cond \mD_n)
	%\propto p(\mD|\nat)p(\nat)
    \propto \exp(- (n_0+n) \bregman(\nat; \MAPt^\text{MAP}))
    \label{eq:joint_likelihood}
\end{align}
reaching its maximum in $\MAPt^\text{MAP}$ such that
\begin{align}
    \nabla \logpart(\MAPt^\text{MAP}) = \MAPm^\text{MAP}
    = \frac{n_0 \meanp_0 + \sum_{i=1}^n T_i}{n_0+n} \; .
\end{align}
where $T_i=T(X_i)$.
When $n_0=0$ -- eg we observed zero samples from the prior -- we recover the Maximum Likelihood Estimate (MLE).
In general, we will write $\MAPt$ to refer to the MAP, and view the MLE as a very special case.
The MLE and MAP estimates are statistics of the dataset $\mD_n$.
Given a random dataset, we wish to bound their deviation from the optimum $\nat^*$ or $\meanp^*$.

@@

\begin{table}[t]
\begingroup
%\renewcommand*{\arraystretch}{1.25}%
\newcommand*{\greencmark}{\textcolor{Green}{\cmark}}
\newcommand*{\redxmark}{\textcolor{Red}{\xmark}}
\centering
\caption{Comparison of Analysis of SMD, \\
under relative smoothness and relative strong convexity assumptions. 
Barrier tells whether the assumption can hold with barrier objectives.
$\lr_n \sim \inv{n}$ says whether the convergence rate hold with this step-size scheduling, and the last column tells whether the rate applies to the last iterate or to some average iterate.
}
\begin{tabular}{lccc}
%{p{.15\textwidth}m{.05\textwidth}m{.08\textwidth}m{.1\textwidth}}
\toprule
Boundedness & Barriers &  $\lr_n \sim \inv{n}$ & Last iterate \\
\midrule
%Strongly convex and bounded variance\newline
%\citep[e.g.,][]{??}
%&
%$
%\cB_A(\theta,\theta') \geq \frac{1}{2}\norm{\theta-\theta'}^2,
%\quad
%\expect[x]{\norm{\m - x}}^2\leq\sigma^2
%$
%\\
Var. on $\Theta$~\eqref{eq:hanzely} % \newline \citep{hanzely2018fastest}
& \redxmark & \greencmark  & \redxmark
\\
Var. at $\theta_*$~\eqref{eq:dragomir} %\newline \citep{dragomir2021fast}
& \redxmark & \greencmark  & \greencmark
\\
Opt. gap~\eqref{eq:dorazio} %\newline \citep{dorazio2021stochastic}
& \greencmark & \redxmark & \greencmark
\\
\bottomrule
\end{tabular}
\label{tbl:assumptions}
\endgroup
\end{table}

@@

\paragraph{Density Estimation}
While statistics has a plethora of asymptotic results \citep{vdv1998asymptotic}, there is less work in finite sample bounds.
Since the KL explode to $\infty$ if the model is unable to give weight to some samples, a lot of this work has been devoted to distances, such as $L^2$ \citep{tsybakov2009introduction}, $L^1$ \citep{devroye2001combinatorial} or more recently Hellinger distance \citep{baraud2017new}.
Contrary to the KL, these distance are symmetric and behave smoothly.

>>>>

Exponential families are among the most widely used statistical models.
Many standard random variables fall into this category: Gaussians, categorical, Gamma or Dirichlet.
They are both flexible enough to model a variety of data sources $X$, and easy to describe with some sufficient statistics $T(X) \in \real^d$.
In particular, they are appreciated for their convex log-likelihood
\alignn{
f(\nat) = \E[-\log p(X \cond \nat)] = \logpart(\nat) - \lin{\E[T(X)] , \nat}
}
where $\logpart$ is a convex function and $\nat$ is a parameter vector.
This convexity lays the foundation for generalized linear models \citep{mccullagh1989generalized} 
or variants of principal component analysis \citep{collins2001generalization} among other applications.
Another interest of exponential families is their closed form maximum-likelihood estimate (MLE) obtained via moment matching
\begin{align}
	\nabla \logpart(\hat \nat_n^\text{MLE}) = \frac{\sum_{i=1}^n T(x_i)}{n} 
	\label{eq:defMLE}
\end{align}
Given a conjugate prior, a similar formula holds for the maximum a posteriori (MAP).

@@

\paragraph{Maximum A Posteriori (MAP).} 
\textcolor{red}{\textbf{D.S.: This section is very unclear. }}
Using Bayes rule with~\eqref{eq:def_expfamily}~\&~\eqref{eq:def_prior}, the posterior of $\nat$ given a dataset $\mD_n =(X_1,\dots,X_n)$ is
\begin{align}
	p(\nat \cond \mD_n)
	%\propto p(\mD|\nat)p(\nat)
    \propto \exp(- (n_0+n) \bregman(\nat; \MAPt^\text{MAP}))
    \label{eq:joint_likelihood}
\end{align}
reaching its maximum in $\MAPt^\text{MAP}$ such that
\begin{align}
    \nabla \logpart(\MAPt^\text{MAP}) = \MAPm^\text{MAP}
    = \frac{n_0 \meanp_0 + \sum_{i=1}^n T_i}{n_0+n} \; .
    \label{eq:defMAP}
\end{align}
where $T_i=T(X_i)$.
When $n_0=0$ -- eg we observed zero samples from the prior -- we recover the Maximum Likelihood Estimate (MLE).
In general, we will write $\MAPt$ to refer to the MAP, and view the MLE as a very special case.
The MLE and MAP estimates are statistics of the dataset $\mD_n$.
Given a random dataset, we wish to bound their deviation from the optimum $\nat^*$ or $\meanp^*$.

@@

\begin{table}[t]
\begingroup
%\renewcommand*{\arraystretch}{1.25}%
\newcommand*{\greencmark}{\textcolor{Green}{\cmark}}
\newcommand*{\redxmark}{\textcolor{Red}{\xmark}}
\centering
\caption{Summary of results for SMD
under relative smoothness and relative strong convexity assumptions. 
Barrier tells whether the assumption can hold with barrier objectives.
$\lr_n \sim \inv{n}$ says whether the convergence rate hold with this step-size scheduling, and the last column tells whether the rate applies to the last iterate or to some average iterate.
}
\begin{tabular}{lccc}
%{p{.15\textwidth}m{.05\textwidth}m{.08\textwidth}m{.1\textwidth}}
\toprule
Boundedness & Barriers &  $\lr_n \sim \inv{n}$ & Last iterate \\
\midrule
%Strongly convex and bounded variance\newline
%\citep[e.g.,][]{??}
%&
%$
%\cB_A(\theta,\theta') \geq \frac{1}{2}\norm{\theta-\theta'}^2,
%\quad
%\expect[x]{\norm{\m - x}}^2\leq\sigma^2
%$
%\\
Var. on $\Theta$~\eqref{eq:hanzely} % \newline \citep{hanzely2018fastest}
& \redxmark & \greencmark  & \redxmark
\\
Var. at $\theta_*$~\eqref{eq:dragomir} %\newline \citep{dragomir2021fast}
& \redxmark & \greencmark  & \greencmark
\\
Opt. gap~\eqref{eq:dorazio} %\newline \citep{dorazio2021stochastic}
& \greencmark & \redxmark & \greencmark
\\
\bottomrule
\end{tabular}
\label{tbl:assumptions}
\endgroup
\end{table}

@@

\paragraph{Parameter Estimation.}
Statistics has a plethora of asymptotic results \citep{vdv1998asymptotic}.
In particular for sum of independent variables such as~\eqref{eq:defMLE}, large deviations theory \citep{varadhan1984large} characterizes the decay of the tail.

\paragraph{Density Estimation.}
Since the KL explode to $\infty$ if the model is unable to give weight to some samples, a lot of this work has been devoted to distances, such as $L^2$ \citep{tsybakov2009introduction}, $L^1$ \citep{devroye2001combinatorial} or more recently Hellinger distance \citep{baraud2017new}.
Contrary to the KL, these distance are symmetric and behave smoothly.
